Understanding meta-analysis through data simulation with application to power analysis

### 2023-07-18 documents/supplementary/supplementary.Rmd
fixing typos and revision

---
title: "Understanding meta-analysis through data simulation with application to power analysis"
subtitle: "Supplementary materials"
bibliography: "../files/references.bib"
csl: "../files/apa.csl"
author:
  - Filippo Gambarota
  - Gianmarco Altoè
output:
  bookdown::pdf_document2:
    toc: true
    latex_engine: xelatex
header-includes:
  - \setcounter{table}{0} 
  - \renewcommand*{\thetable}{S\arabic{table}}
  - \setcounter{figure}{0} 
  - \renewcommand*{\thefigure}{S\arabic{figure}}
  - \setcounter{equation}{0}
  - \renewcommand{\theequation}{S\arabic{equation}}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
---

@@

```{r packages, include = FALSE}
library(dplyr)
library(tidyr)
library(here)
devtools::load_all() # loading all custom functions
```

@@

In the main paper, we presented the basic meta-analysis models and simulations. However, in real-world meta-analyses, it is common to find complex data structures such as multilevel and multivariate. As reported in the paper, there could be a different source of dependencies when each study reports multiple effect sizes. Rephrasing the [@Cheung2015-bw, pp. 121--122] taxonomy, the first type of dependency, which we call *multilevel* dependency, arises when multiple independent experiments are associated with the same study. Level 1 (i.e., participants' level) is different, but the effects are likely correlated because the same research group conducted the experiments. The second type of dependency, *multivariate* dependency, arises when multiple effects are collected on the same pool of participants within each study. Now sampling errors within each study are correlated because the same participant is tested multiple times. Finally, we have a dependency on multiple effects at the population level. Two outcomes could be correlated (e.g., the type of
memory tasks) even if the pool of participants is different. The first two types of dependency are the reasons to use a multilevel and multivariate model. The following sections will show two examples to simulate these complex data structures.

@@

In the paper, we used the `mapply()` function, a particularly useful tool in R, to apply the same function (in parallel) with multiple parameters using multiple vectors. In our case, the `sim_study()` function has several parameters (`theta`, `nt`, among others), and we can apply the function multiple times with a combination of parameters. It can be considered essentially a `for` loop but more compact and less verbose.

@@

# preallocate for speed (not necessary with mapply)
res <- vector(mode = "list", length = nrow(sim))
for(i in 1:length(res)){
  res[[i]] <- sim_study(theta = sim$theta[i], nc = sim$nc[i], nt = sim$nt[i])
}

# using mapply
sim <- mapply(sim_study, sim$theta, sim$nc, sim$nt)

# using sim_studies (with mapply inside)
sim <- sim_studies(sim$theta, sim$nc, sim$nt, data = sim)
```

@@

\begin{align}
\begin{gathered}
d_{ij} = \theta_r + \theta_i + \theta_{ij} + \epsilon_{ij} \\
\theta_i \sim N(0, \tau^2) \\
\theta_{ij} \sim N(0, \omega^2) \\
\epsilon_{ij} \sim N(0,\sigma^2_i)
(\#eq:three-level-model)
\end{gathered}
\end{align}

The notation suggests that each effect size belongs to an experiment (effect size) ($j$) nested within a study ($i$). The variability between studies (level 3) is handled by the $\tau^{2}$ parameter and the variability within studies (level 2) is handled by the $\omega^{2}$ parameter. Practically, each study is composed of the average effect size $\theta_{r}$ plus a study-specific adjustment determined by $\tau^{2}$. Then each effect size is composed of the study-specific
effect ($\theta_{r} + \theta_{i}$) plus the adjustment determined by $\omega^{2}$ thus $\theta_{r} + \theta_{i} + \theta_{\text{ij}}$. The number of effect sizes within each study will determine the estimation precision of the $\omega^{2}$ parameter and the number of studies will determine the estimation precision of the $\tau^{2}$ parameter. The
relationship between $\tau^{2}$ and $\omega^{2}$ can be expressed using the intraclass correlation coefficient (ICC) expressed as $ICC = \frac{\tau^{2}}{\tau^{2} + \omega^{2}}$ representing the proportion of total heterogeneity associated with level 2 or level 3. For example, an ICC of 0.5 suggests that 50% of heterogeneity is caused by between-studies variability. On the extreme, an ICC close to 1 suggests that the between-studies variability causes most heterogeneity
(i.e., the effect sizes within each study are very similar).

```{r sim-multilevel}
set.seed(seed)
i <- 50 # Number of studies (level 3)
j <- 5 # Number of effect sizes within each study (level 2)
tau2 <- 0.3 # heterogeneity between studies
omega2 <- 0.1 # heterogeneity within studies
icc <- tau2 / (tau2 + omega2) # real ICC
n <- 30 # sample size for each group, within each study
theta_r <- 0.3 # real average effect size

theta_i <- rnorm(i, 0, sqrt(tau2))
theta_ij <- rnorm(i*j, 0, sqrt(omega2))

sim <- tidyr::expand_grid(
  study = 1:i,
  effect = 1:j,
  nt = n,
  nc = n,
  theta_r = theta_r
)

sim$theta_i <- theta_i[sim$study]
sim$theta_ij <- theta_ij

@@

Each study $i$ is repeated $j$ times^[To create a more realistic simulation, the number of effect sizes for each paper could be heterogeneous (e.g., sampled from a Poisson distribution)] along the $\theta_{i}$ while each effect has a unique $\theta_{\text{ij}}$. We can use the same approach as the main paper using the `sim_studies()` function.

```{r}
set.seed(seed)
sim$theta_r_i_ij <- sim$theta_r + sim$theta_i + sim$theta_ij
sim <- sim_studies(theta = sim$theta_r_i_ij, nt = sim$nt, nc = sim$nc, data = sim)
```

@@

As explained above, the model estimates the average effect $\theta_{r}$ and the two heterogeneity parameters. To create a more realistic simulation, we can create some variability in the number of experiments within the same paper, similar to what we did for the number of participants in the main paper. We simulated a three-level model, but it is possible to extend to *n* levels for a more complicated nested structure. In this case, we need more heterogeneity parameters (i.e., more nested "adjustments"), but the overall simulation setup is the same.

@@

Following the example of the main paper, we have three memory tests collected on the same pool of participants. For this reason we have three real effects ($\theta_{r_{1}}$, $\theta_{r_{2}}$ and $\theta_{r_{3}}$), three heterogeneity ($\tau^2_1$, $\tau^2_2$ and $\tau^2_3$) and the population-level correlations between these outcomes ($\rho_{12}$, $\rho_{13}$ and $\rho_{23}$). Moreover, the sampling errors ($\epsilon_i$) are now correlated; thus each study will have three variances for each outcome ($\sigma^2_1$, $\sigma^2_2$, and $\sigma^2_3$), and the correlations between sampling errors ($\rho_{s12}$, $\rho_{s13}$ and $\rho_{s23}$)^[We used the subscript $s$ for sampling errors to distinguish between between-outcomes correlations and correlations between sampling errors]. Equation \@ref(eq:multivariate-model) formalize the multivariate model for a single study $i$ with three outcomes. The random-effects adjustments ($\theta_i$) to the overall effect and sampling errors are now sampled from a multivariate normal distribution, respectively $\theta_i \sim \mathcal{MVN}(0, \mathrm{T}^2)$ and $\theta_i \sim \mathcal{MVN}(0, \mathrm{V}^2)$.

\begin{align}
\begin{gathered}
\begin{bmatrix}
d_{i1} \\
d_{i2} \\
d_{i3}
\end{bmatrix} 
=
\begin{bmatrix}
\theta_{1} \\
\theta_{2} \\
\theta_{3}
\end{bmatrix}
+
\begin{bmatrix}
\theta_{i1} \\
\theta_{i2} \\
\theta_{i3}
\end{bmatrix}
+
\begin{bmatrix}
\epsilon_{i1} \\
\epsilon_{i2} \\
\epsilon_{i3}
\end{bmatrix} \\
\begin{bmatrix}
\theta_{i1} \\
\theta_{i2} \\
\theta_{i3}
\end{bmatrix}
\sim \mathcal{MVN}(0, \mathrm{T}^2) \\
\begin{bmatrix}
\epsilon_{i1} \\
\epsilon_{i2} \\
\epsilon_{i3}
\end{bmatrix}
\sim \mathcal{MVN}(0, \mathrm{V}^2) \\
\mathrm{T}^2 = \begin{bmatrix} 
\tau_1^2 & & & \\ 
\rho_{21}\tau_2\tau_1 & \tau_2^2 & & \\ 
\rho_{31}\tau_3\tau_1 & \rho_{32}\tau_3\tau_2 & \tau_3^2
\end{bmatrix} \\
\mathrm{V}^2 = \begin{bmatrix} 
\sigma_1^2 & & & \\ 
\rho_{s21}\sigma_2\sigma_1 & \sigma_2^2 & & \\ 
\rho_{s31}\sigma_3\sigma_1 & \rho_{s32}\sigma_3\sigma_2 & \sigma_3^2
\end{bmatrix} \\
(\#eq:multivariate-model)
\end{gathered}
\end{align}

@@

The previous code simulated a *compound symmetry* (CS) structure where all variances ($\tau^2$) and covariances ($\rho\tau_1\tau_2$) are the same. There could be different structures of the $T^2$ matrix that are clearly explained in the `metafor::rma.mv()` documentation (see https://wviechtb.github.io/metafor/reference/rma.mv.html#specifying-random-effects). For the sake of simplicity we will use the CS structure. We can use the `sim_T()` function that given the number of studies $k$, outcomes $p$, the vector of $\tau^2$ and the correlation $\rho$ generates the random-effect adjustments^[The function allow to specify heterogeneous $\tau^2$ values where the CS structure assume the same $\tau^2$ for each outcome. Using a fixed $\rho$ and different $\tau^2$ will create what `rma.mv()` calls a *heteroscedastic compound symmetric structure* thus a matrix where the correlation is the same between outcomes and each outcome has a different $\tau^2$ value].

@@

We have all pieces to simulate the multivariate model. We can use the `sim_study_m()` that follows the same idea of the `sim_study()` function but extends to multiple outcomes. We generate a single study with $p$ outcomes where sampling errors are correlated according to a compound symmetry structure^[Also for the $V^2$ matrix we can assume different variance-covariance structures. In this example we are using a CS structure]. In practical terms, we assume that the
correlation at the participant level between multiple outcomes is $r$ and fixed between studies.

@@

```{r}
set.seed(seed)
thetas <- c(0.1, 0, 0.7) # three real average outcomes
n <- 30 # number of participant per study
r <- 0.6 # correlation between sampling errors
sim_study_m(thetas, nt = n, nc = n, r = r)
```

@@

thetas_r <- c(0.1, 0, 1) # population level real effect sizes
taus2 <- c(0.3, 0.3, 0.3) # population level real tau2s

@@

# adding the real effects and the random-effect adjustments
sim$theta_r <- rep(thetas_r, k)
sim$theta_i <- sim_T(k, p, taus2, rho)

# average effect + random-effect adjustment
sim$theta_r_i <- sim$theta_r + sim$theta_i

@@

# simulating the effect sizes for each study
resl <- lapply(siml, function(x) sim_study_m(x$theta_r_i, x$nt, x$nc, r))
res <- dplyr::bind_rows(resl) # combine everything
sim <- cbind(sim, res) # append to the sim dataframe

@@

The output is more complicated compared to previous models. The `Variance Components` is the estimation of the $T^{2}$ matrix. In the two-level random-effects model, we have a single $\tau^{2}$ while in the multivariate model, we have a full variance-covariance matrix. Using the `struct = "CS"` argument, we are forcing the `rma.mv()` function to use a compound symmetry structure, thus estimating a single $\tau^{2}$ and a single $\rho$. Again, this is consistent with our simulation but could be a stringent assumption in a real-world analysis. The crucial part is the `random = ~ p|id` argument that specifies the random-effects structure. The syntax is clearly explained here [https://wviechtb.github.io/metafor/reference/rma.mv.html#specifying-random-effects](https://wviechtb.github.io/metafor/reference/rma.mv.html#specifying-random-effects) The basic idea is to specify the random effect structure as `~ inner|outer` where inner is the outcome variable and outer is the paper variable. The mods = `~ 0 + p` argument specify to estimate the average effect for each outcome (i.e., the $\theta_{r}$ vector of average effects). We are removing the intercept (`~ 0 + p` or equivalently `~ p - 1`) and the model will estimate the mean for each level of the `p` variable.

We could use also another structure for the $T^2$ matrix. We know that our data are generated using a CS structure thus using another structure will be less appropriate. In reality, is not known which is the real data generation model thus we can try different structure. Setting the `struct = "UN"` (unstructured) argument, `rma.mv` try to estimate a variance-covariance matrix where all correlations and variances need to be estimated. Compared to the previous model, we are know estimating 3 heterogeneity and three correlation parameters.

@@

@@

<!-- CAPTIONS -->

@@

>>>>

---
title: "Understanding meta-analysis through data simulation with application to power analysis"
subtitle: "Supplementary materials"
bibliography: "../files/references.bib"
csl: "../files/apa.csl"
author:
  - Filippo Gambarota
  - Gianmarco Altoè
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 1
    latex_engine: xelatex
header-includes:
  - \setcounter{table}{0} 
  - \renewcommand*{\thetable}{S\arabic{table}}
  - \setcounter{figure}{0} 
  - \renewcommand*{\thefigure}{S\arabic{figure}}
  - \setcounter{equation}{0}
  - \renewcommand{\theequation}{S\arabic{equation}}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
  - \newcommand{\rev}[1]{\textcolor{red}{#1}}
---

@@

```{r packages, include = FALSE}
library(dplyr)
library(tidyr)
library(here)
devtools::load_all() # loading all custom functions

# theme for plots
theme_paper <- function(base_size = 15){
  ltxplot::theme_latex(base_size = base_size)
}
```

@@

In the main paper, we presented the most straightforward situation for a meta-analysis when each included study contributes with a single effect size. However, in real-world meta-analyses, it is common to find complex data structures such as multilevel and multivariate. There are numerous sources of dependency for a meta-analysis [@Cheung2014-fg; @Cheung2015-bw, pp. 121-122].

The first type of dependency (*multilevel*) consists of multiple effect sizes from independent participants. Despite being collected with different participants, effect sizes from the same study could be correlated. We obtain a multilevel data structure with independent effect sizes in the same study. Figure \@ref(fig:img-multilevel) depict the general idea of a *multivariate* model. There are multiple ways to model this data structure [see @Cheung2014-fg for an overview]. Still, the most common way is to use a *multilevel meta-analysis model,* considering the heterogeneity between and within studies. The standard *random-effects* (but also the *equal-effects model*) model is also called a *two-level* model because it combines data after aggregating the participants-level (*first-level*) data. The *three-level* model is commonly used to manage the multilevel data structure [@Cheung2014-fg; @Cheung2019-po]. Compared to the *two-level* model, the *three-level* model estimates two heterogeneity parameters ($\tau^{2}$ and $\omega^{2}$), one for the heterogeneity between studies and the other for the heterogeneity within studies.

A second type of dependency (*multivariate*) consists of multiple effect sizes collected from the same pool of participants. Figure \@ref(fig:img-multivariate) depict the general idea of a *multivariate* model. For example, when using two cognitive tasks on the same pool of participants. This is commonly known as *multivariate meta-analysis* because the effect sizes sampling errors are correlated given that participants are assessed multiple times. A *multivariate meta-analysis model* could be used to
consider the correlation between sampling errors.

A third type of dependency arises when true effect sizes are correlated at the population level [@Cheung2015-bw, pp. 121-122]. For example, two different memory tasks could be correlated because the latent psychological constructs are intrinsically correlated. This correlation is present even when other groups of participants perform the two memory tasks (i.e., no *multivariate* dependency) or are administered by different researchers (i.e., no *multilevel* dependency). This type of dependency could be managed by estimating the correlation between outcomes using a *multivariate random-effects model* [@Cheung2015-bw, pp. 127-130]. Figures \@ref(fig:img-multilevel) and \@ref(fig:img-multivariate) depict the conceptual idea of *multilevel* and *multivariate* models. The following sections will show two examples to simulate these complex data structures.

```{r img-multilevel, echo=FALSE, out.width="80%", fig.cap = get_caption()}
knitr::include_graphics("img/multilevel.pdf")
```

```{r img-multivariate, echo=FALSE, out.width="80%", fig.cap = get_caption()}
knitr::include_graphics("img/multivariate.pdf ")
```

@@

In the paper, we used the `mapply()` function, a particularly useful tool in R, to apply the same function (in parallel) with multiple parameters using multiple vectors. In our case, the `sim_study()` function has several parameters (`es`, `nt`, among others), and we can apply the function multiple times with a combination of parameters. It can be considered essentially a `for` loop but more compact and less verbose.

@@

# preallocate for speed (not necessary with mapply)
res <- vector(mode = "list", length = nrow(sim))
for(i in 1:length(res)){
  res[[i]] <- sim_study(theta = sim$es[i], nc = sim$nc[i], nt = sim$nt[i])
}

# using mapply
sim <- mapply(sim_study, sim$es, sim$nc, sim$nt)

# using sim_studies (with mapply inside)
sim <- sim_studies(sim$es, sim$nc, sim$nt, data = sim)
```

@@

\begin{align}
\begin{gathered}
y_{ij} = \mu_{\theta} + \delta_i + \zeta_{ij} + \epsilon_{ij}
(\#eq:three-level-model1)
\end{gathered}
\end{align}

\begin{align}
\begin{gathered}
\delta_i\sim N(0, \tau^2)
(\#eq:three-level-model2)
\end{gathered}
\end{align}

\begin{align}
\begin{gathered}
\zeta_{ij} \sim N(0, \omega^2)
(\#eq:three-level-model3)
\end{gathered}
\end{align}

\begin{align}
\begin{gathered}
\epsilon_{ij} \sim N(0,\sigma_{\epsilon_i}^{2})
(\#eq:three-level-model4)
\end{gathered}
\end{align}

The notation suggests that each effect size belongs to an experiment (effect size) ($j$) nested within a study ($i$). The variability between studies (level 3) is handled by the $\tau^{2}$ parameter (see Equation \@ref(eq:three-level-model2) and the variability within studies (level 2) is handled by the $\omega^{2}$ parameter (see Equation \@ref(eq:three-level-model3)). Practically, each study is composed of the average effect size $\mu_{\theta}$ plus a study-specific adjustment determined by $\tau^{2}$. Then each effect size is composed of the study-specific effect ($\mu_{\theta} + \delta_{i}$) plus the adjustment determined by $\omega^{2}$ thus $\mu_{\theta} + \delta_{i} + \zeta_{ij}$. As in the two-level meta-analysis model there is also the random error component $\sigma_{\epsilon_i}^{2}$ (see Equation \@ref(eq:three-level-model4)). The number of effect sizes within each study will determine the estimation precision of the $\omega^{2}$ parameter and the number of studies will determine the estimation precision of the $\tau^{2}$ parameter. The relationship between $\tau^{2}$ and $\omega^{2}$ can be expressed using the intraclass correlation coefficient (ICC) expressed as $ICC = \frac{\tau^{2}}{\tau^{2} + \omega^{2}}$ representing the proportion of total heterogeneity associated with level 2 or level 3 [see @Cheung2014-fg]. For example, an ICC of 0.5 suggests that 50% of heterogeneity is caused by between-studies variability. On the extreme, an ICC close to 1 suggests that the between-studies variability causes most heterogeneity (i.e., the effect sizes within each study are very similar).

```{r sim-multilevel}
set.seed(seed)
i <- 50 # Number of studies (level 3)
j <- 5 # Number of effect sizes within each study (level 2)
tau2 <- 0.3 # heterogeneity between studies
omega2 <- 0.1 # heterogeneity within studies
icc <- tau2 / (tau2 + omega2) # real ICC
n <- 30 # sample size for each group, within each study
mu <- 0.3 # real average effect size

delta_i <- rnorm(i, 0, sqrt(tau2))
zeta_ij <- rnorm(i*j, 0, sqrt(omega2))

sim <- tidyr::expand_grid(
  study = 1:i,
  effect = 1:j,
  nt = n,
  nc = n,
  mu = mu
)

sim$delta_i <- delta_i[sim$study]
sim$zeta_ij <- zeta_ij

@@

Each study $i$ is repeated $j$ times^[To create a more realistic simulation, the number of effect sizes for each paper could be heterogeneous (e.g., sampled from a Poisson distribution)] along the $\delta_{i}$ while each effect has a unique $\zeta_{\text{ij}}$. We can use the same approach as the main paper using the `sim_studies()` function.

```{r}
set.seed(seed)
sim$es <- sim$mu + sim$delta_i + sim$zeta_ij
sim <- sim_studies(es = sim$es, nt = sim$nt, nc = sim$nc, data = sim)
```

@@

As explained above, the model estimates the average effect $\mu_{\theta}$ and the two heterogeneity parameters. To create a more realistic simulation, we can create some variability in the number of experiments within the same paper, similar to what we did for the number of participants in the main paper. We simulated a three-level model, but it is possible to extend to *n* levels for a more complicated nested structure. In this case, we need more heterogeneity parameters (i.e., more nested "adjustments"), but the overall simulation setup is the same.

@@

Following the example of the main paper, we have three memory tests collected on the same pool of participants. For this reason we have three real effects ($\mu_{\theta_{1}}$, $\mu_{\theta_{2}}$ and $\mu_{\theta_{3}}$), three heterogeneity ($\tau^2_1$, $\tau^2_2$ and $\tau^2_3$) and the population-level correlations between these outcomes ($\rho_{12}$, $\rho_{13}$ and $\rho_{23}$). Moreover, the sampling errors ($\epsilon_i$) are now correlated; thus each study will have three sampling variances for each outcome ($\sigma_{\epsilon_1}^{2}$, $\sigma_{\epsilon_1}^{2}$, and $\sigma_{\epsilon_1}^{2}$), and the correlations between sampling errors ($\rho_{s12}$, $\rho_{s13}$ and $\rho_{s23}$)^[We used the subscript $s$ for sampling errors to distinguish between between-outcomes correlations and correlations between sampling errors]. Equations \@ref(eq:multivariate-model1), \@ref(eq:multivariate-model2), \@ref(eq:multivariate-model3), \@ref(eq:multivariate-model4) \@ref(eq:multivariate-model5) formalize the multivariate model for a single study $i$ with three outcomes. The random-effects adjustments ($\theta_i$) to the overall effect and sampling errors are now sampled from a multivariate normal distribution, respectively $\theta_i \sim \mathcal{MVN}(0, \mathrm{T}^2)$ and $\theta_i \sim \mathcal{MVN}(0, \mathrm{V}^2)$.

\begin{align}
\begin{gathered}
\begin{bmatrix}
y_{1_i} \\
y_{2_i} \\
y_{3_i}
\end{bmatrix} 
=
\begin{bmatrix}
\mu_{{\theta}_{1}} \\
\mu_{{\theta}_{2}} \\
\mu_{{\theta}_{3}}
\end{bmatrix}
+
\begin{bmatrix}
\delta_{1_i} \\
\delta_{2_i} \\
\delta_{3_i}
\end{bmatrix}
+
\begin{bmatrix}
\epsilon_{1_i} \\
\epsilon_{2_i} \\
\epsilon_{3_i}
\end{bmatrix}
(\#eq:multivariate-model1)
\end{gathered}
\end{align}

\begin{align}
\begin{gathered}
\begin{bmatrix}
\delta_{1_i} \\
\delta_{2_i} \\
\delta_{3_i}
\end{bmatrix}
\sim \mathcal{MVN}(0, \mathrm{T})
(\#eq:multivariate-model2)
\end{gathered}
\end{align}

\begin{align}
\begin{gathered}
\begin{bmatrix}
\epsilon_{1_i} \\
\epsilon_{2_i} \\
\epsilon_{3_i}
\end{bmatrix}
\sim \mathcal{MVN}(0, \mathrm{V})
(\#eq:multivariate-model3)
\end{gathered}
\end{align}

\begin{align}
\begin{gathered}
\mathrm{T} = \begin{bmatrix}
\tau_1^2 & & & \\
\rho_{21}\tau_2\tau_1 & \tau_2^2 & & \\
\rho_{31}\tau_3\tau_1 & \rho_{32}\tau_3\tau_2 & \tau_3^2
\end{bmatrix}
(\#eq:multivariate-model4)
\end{gathered}
\end{align}

\begin{align}
\begin{gathered}
\mathrm{V} = \begin{bmatrix}
\sigma^2_{\epsilon_1} & & & \\
\rho_{s_{21}}\sigma_{\epsilon_2}\sigma_{\epsilon_1} & \sigma^2_{\epsilon_2} & & \\
\rho_{s_{31}}\sigma_{\epsilon_3}\sigma_{\epsilon_1} & \rho_{s_{32}}\sigma_{\epsilon_3}\sigma_{\epsilon_2} & \sigma^2_{\epsilon_3}
\end{bmatrix}
(\#eq:multivariate-model5)
\end{gathered}
\end{align}

@@

The previous code simulated a *compound symmetry* (CS) structure where all variances ($\tau^2$) and covariances ($\rho\tau_1\tau_2$) are the same. There could be different structures of the $T$ matrix that are clearly explained in the `metafor::rma.mv()` documentation (see https://wviechtb.github.io/metafor/reference/rma.mv.html#specifying-random-effects). For the sake of simplicity we will use the CS structure. We can use the `sim_T()` function that given the number of studies $k$, outcomes $p$, the vector of $\tau^2$ and the correlation $\rho$ generates the random-effect adjustments^[The function allow to specify heterogeneous $\tau^2$ values where the CS structure assume the same $\tau^2$ for each outcome. Using a fixed $\rho$ and different $\tau^2$ will create what `rma.mv()` calls a *heteroscedastic compound symmetric structure* thus a matrix where the correlation is the same between outcomes and each outcome has a different $\tau^2$ value].

@@

We have all pieces to simulate the multivariate model. We can use the `sim_study_m()` that follows the same idea of the `sim_study()` function but extends to multiple outcomes. We generate a single study with $p$ outcomes where sampling errors are correlated according to a compound symmetry structure^[Also for the $V$ matrix we can assume different variance-covariance structures. In this example we are using a CS structure]. In practical terms, we assume that the
correlation at the participant level between multiple outcomes is $r$ and fixed between studies.

@@

```{r}
set.seed(seed)
mus <- c(0.1, 0, 0.7) # three real average outcomes
n <- 30 # number of participant per study
r <- 0.6 # correlation between sampling errors
sim_study_m(mus, nt = n, nc = n, r = r)
```

@@

mus <- c(0.1, 0, 1) # population level real effect sizes
taus2 <- c(0.3, 0.3, 0.3) # population level real tau2s

@@

# adding the real effects and the random-effect adjustments
sim$mu <- rep(mus, k)
sim$delta_i <- sim_T(k, p, taus2, rho)

# average effect + random-effect adjustment
sim$es <- sim$mu + sim$delta_i

@@

# simulating the effect sizes for each study
resl <- lapply(siml, function(x) sim_study_m(x$es, x$nt, x$nc, r))
res <- dplyr::bind_rows(resl) # combine everything
sim <- cbind(sim, res) # append to the sim dataframe

@@

The output is more complicated compared to previous models. The `Variance Components` is the estimation of the $T^{2}$ matrix. In the two-level random-effects model, we have a single $\tau^{2}$ while in the multivariate model, we have a full variance-covariance matrix. Using the `struct = "CS"` argument, we are forcing the `rma.mv()` function to use a compound symmetry structure, thus estimating a single $\tau^{2}$ and a single $\rho$. Again, this is consistent with our simulation but could be a stringent assumption in a real-world analysis. The crucial part is the `random = ~ p|id` argument that specifies the random-effects structure. The syntax is clearly explained here [https://wviechtb.github.io/metafor/reference/rma.mv.html#specifying-random-effects](https://wviechtb.github.io/metafor/reference/rma.mv.html#specifying-random-effects) The basic idea is to specify the random effect structure as `~ inner|outer` where inner is the outcome variable and outer is the paper variable. The mods = `~ 0 + p` argument specify to estimate the average effect for each outcome (i.e., the $\mathbf{\mu_{\theta}}$ vector of average effects). We are removing the intercept (`~ 0 + p` or equivalently `~ p - 1`) and the model will estimate the mean for each level of the `p` variable.

We could use also another structure for the $T$ matrix. We know that our data are generated using a CS structure thus using another structure will be less appropriate. In reality, is not known which is the real data generation model thus we can try different structure. Setting the `struct = "UN"` (unstructured) argument, `rma.mv` try to estimate a variance-covariance matrix where all correlations and variances need to be estimated. Compared to the previous model, we are know estimating 3 heterogeneity and three correlation parameters.

@@

# Extra

## $I^2$ with homogeneous vs heterogeneous sample sizes

In the paper we suggested that is not straightforward to simulate a meta-analysis fixing the $I^2$ as a simulation parameter with heterogeneous sample sizes. Here we show that when sample sizes are generated from a probability distribution (e.g., Gaussian) we can fix the $I^2$ value using the $\tilde{v}$ calculated using the expected value of the sample size distribution. For example, if the sample sizes of our $k$ studies is generate from a Poisson distribution $n_i \sim \mathcal{P}(\lambda)$, then $\tilde{v}$ can be calculated using $\lambda$ as *typical* sample size (i.e., as if the sample sizes were homogeneous). The following code simulate this scenario showing the distribution of $I^2$ with heterogeneous sample sizes^[Compared to the simulations of the paper here we are simulating the effect size and the sampling variances from the respective sampling distributions [e.g., supplementary materials by @Van_Aert2019-xt]].

```{r}
set.seed(seed)
tau2 <- 0.2 # real tau2 value
navg <- 40 # average sample size
nmin <- 10 # minimum sample size
k <- 30 # number of studies
k <- 30 # number of studies
nsim <- 1000 # number of simulations
mu <- 0.3 # real average effect size

# here we simulate some effect sizes using the aggregated
# data approach, thus from the summary statistics distributions

I2 <- rep(0, nsim)

for(i in 1:nsim){
  ni <- nmin + rpois(k, navg - nmin)

  # effect sizes
  yi <- rnorm(k, rnorm(k, mu, sqrt(tau2)), sqrt(1/ni + 1/ni))

  # sampling variances
  vi <- (rchisq(k, ni + ni - 2) / (ni + ni - 2)) * (1/ni + 1/ni)

  # model
  fit <- rma(yi, vi)

  # extracting I2
  I2[i] <- fit$I2
}
```

```{r, echo = FALSE}
title <- sprintf("Simulated $I^2$ values (Mean = %.3f)", mean(I2))

hist(I2, 
     breaks = seq(min(I2), 100, 3),
     xlab = latex2exp::TeX("$I^2$"),
     main = latex2exp::TeX(title),
     col = "salmon",
     cex.lab = 1.1,
     cex.main = 1.3)
```

The $I^2$ calculated using the expected value of the sample sizes distribution ($\lambda = 40$) and assuming sample size homogeneity is:

```{r}
v <- 1/navg + 1/navg
tau2 / (v + tau2)
```

The value is very close to the mean of the simulated $I^2$ distribution. In conclusion, by generating the sample sizes from a probability distribution (e.g., Poisson) and using the Higgins and Thompson [-@Higgins2002-fh] formula to calculate $\tilde{v}$ we can somehow fix the $I^2$ with heterogeneous sample sizes using the expected value of the distribution.

@@

@@

(ref:img-multilevel) Graphical representation of a *multilevel* (three-level) model. Each observed effect size is defined as the overall effect average effect $\mu_{\theta}$, a study-specific random-effect $\delta_{i} \sim N(0, \tau^2)$ and a experiment-specific random-effects $\zeta_{ij} N(0, \omega^2)$.

(ref:img-multivariate) Graphical representation of a *multivariate* model. In this example, we have two outcomes ($p$) (magenta and green) with an average effect ($\mu_{\theta_p}$), heterogeneity parameters ($\tau^2_p$), and a certain true correlation value ($\rho_{p_1p_2}$, third type of dependency). Each study at level-2 (as in the standard random-effects model) has two effect sizes with corresponding sampling variances. The two effect sizes are correlated ($\rho_{s}$, with $s$ for sampling errors) because each effect is collected on the same participants.
